---
title: "Assignment 3A: Global Baseline Estimate"
author: "Andre Thomson"
format:
  html:
    toc: true
    embed-resources: true
    code-fold: true
    code-summary: "Show the code"
    code-tools: true
editor: visual
execute:
  warning: false
  message: false
---

## Approach

I used the movie ratings from my Week 2 survey. There are six participants and six movies. Everyone rated all six movies, so there was no missing rating to estimate.

To test the Global Baseline Estimate, I held out the last rating in the file. That rating belongs to Participant 6 for *The Super Mario Bros. Movie*. I used the other 35 ratings to make the estimate. The original CSV stays unchanged.

## Load the packages and data

This code loads the CSV and checks the three columns I need.

```{r}
#| label: load-data

library(tidyverse)
library(knitr)

# Read the ratings exported from the Week 2 database.
ratings <- read_csv(
  "movie_ratings_query.csv",
  show_col_types = FALSE
)

# Make sure the expected columns are present.
required_columns <- c(
  "participant_id",
  "movie_title",
  "rating"
)

stopifnot(
  all(required_columns %in% names(ratings)),
  all(ratings$rating %in% 1:5),
  !anyDuplicated(ratings[c("participant_id", "movie_title")])
)

glimpse(ratings)
```

## Check the ratings

I checked the number of participants, movies, and ratings before doing the calculation.

```{r}
#| label: check-data

# Count the records and make sure the Week 2 data is complete.
data_check <- ratings |>
  summarise(
    participants = n_distinct(participant_id),
    movies = n_distinct(movie_title),
    ratings = n(),
    missing_ratings = sum(is.na(rating))
  )

stopifnot(
  data_check$participants == 6,
  data_check$movies == 6,
  data_check$ratings == 36,
  data_check$missing_ratings == 0
)

kable(data_check)
```

## Set aside one rating

The course spreadsheet estimates a question-mark value. My survey has no blanks, so I set aside the last rating for this demonstration.

```{r}
#| label: hold-out-rating

target_user <- "Participant 6"
target_movie <- "The Super Mario Bros. Movie"

# Save the real value so it can be checked after the estimate.
held_out_rating <- ratings |>
  filter(
    participant_id == target_user,
    movie_title == target_movie
  )

stopifnot(nrow(held_out_rating) == 1)

# Remove only the target row from the calculation data.
training_ratings <- ratings |>
  anti_join(
    held_out_rating,
    by = c("participant_id", "movie_title", "rating")
  )

stopifnot(nrow(training_ratings) == 35)

kable(held_out_rating)
```

## Calculate the global average

The global average is the mean of the 35 ratings used in the estimate.

```{r}
#| label: global-average

global_average <- mean(training_ratings$rating)

tibble(
  calculation = "Global average",
  value = global_average
) |>
  kable(digits = 3)
```

## Calculate the participant bias

Participant bias shows whether the participant normally rates above or below the global average.

```{r}
#| label: participant-bias

participant_summary <- training_ratings |>
  group_by(participant_id) |>
  summarise(
    participant_average = mean(rating),
    .groups = "drop"
  ) |>
  mutate(
    participant_bias = participant_average - global_average
  )

target_participant <- participant_summary |>
  filter(participant_id == target_user)

kable(target_participant, digits = 3)
```

## Calculate the movie bias

Movie bias shows whether the movie is normally rated above or below the global average.

```{r}
#| label: movie-bias

movie_summary <- training_ratings |>
  group_by(movie_title) |>
  summarise(
    movie_average = mean(rating),
    .groups = "drop"
  ) |>
  mutate(
    movie_bias = movie_average - global_average
  ) |>
  arrange(desc(movie_average))

target_movie_summary <- movie_summary |>
  filter(movie_title == target_movie)

kable(target_movie_summary, digits = 3)
```

## Estimate the missing rating

I used the same formula shown in the course spreadsheet:

\[
\text{estimate} = \text{global average} + \text{participant bias} + \text{movie bias}
\]

```{r}
#| label: baseline-estimate

# Add the three parts of the Global Baseline Estimate.
baseline_estimate <- global_average +
  target_participant$participant_bias +
  target_movie_summary$movie_bias

# Put the calculation into a small table.
estimate_result <- tibble(
  participant = target_user,
  movie = target_movie,
  global_average = global_average,
  participant_bias = target_participant$participant_bias,
  movie_bias = target_movie_summary$movie_bias,
  estimated_rating = baseline_estimate,
  actual_rating = held_out_rating$rating,
  difference = held_out_rating$rating - baseline_estimate
)

# These values independently check the calculation used in this report.
stopifnot(
  isTRUE(all.equal(global_average, 3.6)),
  isTRUE(all.equal(baseline_estimate, 4.6))
)

kable(estimate_result, digits = 2)
```

The estimated rating is `r sprintf("%.2f", baseline_estimate)`. The original rating was `r held_out_rating$rating`, so the estimate is `r sprintf("%.2f", abs(held_out_rating$rating - baseline_estimate))` points away from the actual rating.

## Recommendation

```{r}
#| label: recommendation

# The held-out movie is the only unrated movie in the demonstration data.
recommendation <- tibble(
  participant = target_user,
  recommended_movie = target_movie,
  estimated_rating = baseline_estimate
)

kable(recommendation, digits = 2)
```

For this demonstration, I would recommend *The Super Mario Bros. Movie* to Participant 6 because its estimated rating is 4.60 out of 5.

This is a small example. Since the original survey has only six people and everyone rated every movie, it is not enough for a real recommendation service.

## Tools and AI assistance

I used R and Quarto for the calculations and report. I used GitHub Copilot in Visual Studio Code to check the code structure, comments, and file names. I checked the numbers by running the calculation separately. Copilot was not the source of the ratings or the final result.

## Conclusion

The Global Baseline Estimate used the overall average, Participant 6's rating pattern, and the movie's rating pattern. It estimated a rating of 4.60. The held-out rating was 5.

## Sources

- CUNY School of Professional Studies. *MovieRatings.xlsx*. DATA 607 course spreadsheet, 2026.
- Week 2 movie ratings collected through my Google Form.
- GitHub. *GitHub Copilot documentation*. <https://docs.github.com/en/copilot>
- R documentation. *mean*. <https://stat.ethz.ch/R-manual/R-devel/library/base/html/mean.html>
- dplyr documentation. *summarise*. <https://dplyr.tidyverse.org/reference/summarise.html>
