---
title: "Week 1 Assignment"
author: "Andre Thomson"
output:
  html_document:
    toc: true
    self_contained: true
---

## Approach

For this assignment, I used a prostate cancer dataset from Kaggle. The dataset has 97 observations and 10 variables. Some of the variables include age, prostate weight (the weight of the prostate gland), cancer volume (the amount of cancer found in the prostate tissue), Gleason score (a grading system used to describe how abnormal prostate cancer cells look), seminal vesicle invasion (whether the cancer has spread into the seminal vesicles), PSA (prostate-specific antigen, a protein measured in the blood), and a training indicator (shows whether an observation was part of the original training set or test set used for statistical modeling).

I chose this dataset because prostate cancer affects many men, and I am interested in health data. I wanted to use this dataset to look at some of the variables connected to prostate cancer.

For this assignment, I will load the data into R, check the structure, look for missing values and duplicates, select the variables I want to focus on, rename the abbreviated columns, and create a few simple charts. I will use the results to see what stands out in the data.

**Source:** The dataset is from Kaggle:

<https://www.kaggle.com/datasets/soujanyahp/prostate-cancer-dataset>

## Research Question

What relationships can be seen between age, cancer volume, Gleason score, seminal vesicle invasion, and PSA in this prostate cancer dataset?

To help answer this question, I will also look at:

- Is log cancer volume related to log PSA?
- How are the observations distributed across Gleason score groups?
- How does seminal vesicle invasion status differ across Gleason score groups?
- How are the observations distributed across age groups?


## Load the data

I loaded the `tidyverse` package and connected the assignment to the prostate CSV file. The code checks the assignment folder and the Week01 folder for the file.

```{r load-packages, message=FALSE, warning=FALSE}
# Load the tidyverse package for importing, changing, and plotting the data.
library(tidyverse)
```

```{r load-prostate-data}
# Use the CSV file in the same folder as this assignment.
# Read the prostate cancer CSV file from my GitHub repository.
# Check where the CSV file is located.
if (file.exists("prostate.csv")) {
  data_file <- "prostate.csv"
} else if (file.exists("Week01/prostate.csv")) {
  data_file <- "Week01/prostate.csv"
} else {
  stop("prostate.csv was not found. Put the CSV file in the assignment folder.")
}

# Load the prostate cancer dataset.
prostate <- read_csv(
  data_file,
  show_col_types = FALSE
)
```

## Check the data

I checked the first few rows, the size of the dataset, the column names, and the data types.

```{r inspect-prostate-data}
# Look at the first few rows of the dataset.
head(prostate)
# Check the number of rows and columns.
dim(prostate)
# View the column names.
names(prostate)
# Check the structure and data types.
glimpse(prostate)
```

The dataset has `r nrow(prostate)` rows and `r ncol(prostate)` columns.

## Check for missing values and duplicates

I checked for missing values and duplicate rows before changing the data.

```{r check-missing-and-duplicates}
# Count missing values in each column.
missing_values <- colSums(is.na(prostate))
total_missing <- sum(is.na(prostate))
# Count duplicate rows.
duplicate_rows <- sum(duplicated(prostate))

missing_values
cat("Total missing values:", total_missing, "\n")
cat("Duplicate rows:", duplicate_rows, "\n")
```

The dataset has `r total_missing` missing values and `r duplicate_rows` duplicate rows.

## Select and rename variables

I selected the variables I wanted to use and renamed the abbreviated columns so they are easier to read.

```{r select-and-rename-variables}
# Keep the variables I want to use for this assignment.
prostate_selected <- prostate |>
  select(
    lcavol,
    lweight,
    age,
    svi,
    gleason,
    lpsa
  ) |>
  # Rename abbreviated columns so they are easier to understand.
  rename(
    log_cancer_volume = lcavol,
    log_prostate_weight = lweight,
    seminal_vesicle_invasion = svi,
    gleason_score = gleason,
    log_psa = lpsa
  )

head(prostate_selected)
```

## Basic summary

I created a short summary of the dataset.

```{r summarize-prostate-data}
# Create a small summary of the selected variables.
summary_results <- prostate_selected |>
  summarise(
    observations = n(),
    average_age = mean(age, na.rm = TRUE),
    minimum_age = min(age, na.rm = TRUE),
    maximum_age = max(age, na.rm = TRUE),
    average_log_psa = mean(log_psa, na.rm = TRUE)
  )

# Display the summary as a simple table.
knitr::kable(
  summary_results |>
    mutate(across(where(is.numeric), ~ round(.x, 2))),
  caption = "Basic summary of the prostate dataset"
)
```

## Chart Color Key

I used the same colors across the charts to make them easier to follow. Blue is used for the No group, green and orange are used for the age groups, and red is used for the Yes group or the trend line. The grouped bar chart uses blue and red to compare invasion status. The scatter graph uses blue points and a red trend line.

## Bar chart: Gleason scores

This bar chart shows how many observations are in each Gleason score group.

```{r plot-gleason-score-counts}
# Count how many observations are in each Gleason score group.
gleason_counts <- prostate_selected |>
  count(gleason_score)

# Create a bar chart of Gleason score counts.
ggplot(
  gleason_counts,
  aes(
    x = factor(gleason_score),
    y = n,
    fill = factor(gleason_score)
  )
) +
  geom_col() +
  geom_text(
    aes(label = n),
    vjust = -0.3
  ) +
  scale_fill_manual(
    values = c(
      "6" = "#4E79A7",
      "7" = "#59A14F",
      "8" = "#F28E2B",
      "9" = "#E15759"
    ),
    guide = "none"
  ) +
  labs(
    title = "Number of Observations by Gleason Score",
    x = "Gleason Score",
    y = "Count"
  ) +
  theme_minimal()
```

## Barplot: age groups

I grouped the ages into three categories so the counts are easy to compare.

```{r plot-age-group-counts}
# Group ages into three categories.
age_groups <- cut(
  prostate_selected$age,
  breaks = c(-Inf, 59, 69, Inf),
  labels = c("Under 60", "60 to 69", "70 and older")
)

# Count the observations in each age group.
age_counts <- table(age_groups)

# Create the age group barplot.
age_positions <- barplot(
  age_counts,
  main = "Number of Observations by Age Group",
  xlab = "Age Group",
  ylab = "Count",
  col = c("#4E79A7", "#59A14F", "#F28E2B")
)

text(
  age_positions,
  age_counts,
  labels = age_counts,
  pos = 3
)
```

## Grouped bar chart: Gleason score and seminal vesicle invasion

This grouped bar chart compares Gleason score groups by seminal vesicle invasion status.

```{r plot-gleason-by-svi}
# Create counts for Gleason score and seminal vesicle invasion status.
svi_gleason_counts <- prostate_selected |>
  mutate(
    svi_status = factor(
      seminal_vesicle_invasion,
      levels = c(0, 1),
      labels = c("No", "Yes")
    )
  ) |>
  count(gleason_score, svi_status)

# Create a grouped bar chart to compare the two invasion groups.
ggplot(
  svi_gleason_counts,
  aes(
    x = factor(gleason_score),
    y = n,
    fill = svi_status
  )
) +
  geom_col(position = "dodge") +
  scale_fill_manual(
    values = c(
      "No" = "#4E79A7",
      "Yes" = "#E15759"
    )
  ) +
  labs(
    title = "Gleason Score by Seminal Vesicle Invasion",
    x = "Gleason Score",
    y = "Count",
    fill = "Seminal Vesicle Invasion"
  ) +
  theme_minimal()
```

## Graph: cancer volume and PSA

I used a simple scatter graph to look at the relationship between log cancer volume and log PSA.

```{r calculate-cancer-volume-psa-correlation}
# Calculate the correlation between log cancer volume and log PSA.
cancer_volume_correlation <- cor(
  prostate_selected$log_cancer_volume,
  prostate_selected$log_psa,
  use = "complete.obs"
)

round(cancer_volume_correlation, 3)
```

```{r plot-cancer-volume-vs-psa}
# Create a scatter plot of log cancer volume and log PSA.
plot(
  prostate_selected$log_cancer_volume,
  prostate_selected$log_psa,
  main = "Log Cancer Volume and Log PSA",
  xlab = "Log Cancer Volume",
  ylab = "Log PSA",
  pch = 19,
  col = "#4E79A7"
)

# Add a trend line to make the relationship easier to see.
abline(
  lm(
    log_psa ~ log_cancer_volume,
    data = prostate_selected
  ),
  col = "#E15759",
  lwd = 2
)
```

The correlation is about `r round(cancer_volume_correlation, 2)`. The graph shows a positive relationship between log cancer volume and log PSA in this dataset.

## Findings

The main finding is the relationship between log cancer volume and log PSA. The correlation is about `r round(cancer_volume_correlation, 2)`, which shows a positive relationship in this dataset.

The Gleason score chart shows that the observations are not evenly divided across the score groups.

The seminal vesicle invasion chart shows that most observations are in the No group and fewer are in the Yes group.

The age chart shows how the observations are divided across the three age groups.

Based on these results, log cancer volume and log PSA are related in this dataset. The other charts show how the observations are grouped by Gleason score, seminal vesicle invasion, and age. These results describe this dataset only and do not show cause and effect.

## Conclusion

This assignment gave me practice loading a dataset into R, checking the data, selecting variables, renaming columns, and creating simple charts.

The strongest result was the positive relationship between log cancer volume and log PSA. The charts also showed how the observations were grouped by Gleason score, age, and seminal vesicle invasion.

To extend this assignment, I would use a larger prostate cancer dataset and compare more variables with PSA. I would also check another dataset to see if the same patterns appear.

## References

Kaggle. (n.d.). *Prostate cancer dataset*. https://www.kaggle.com/datasets/soujanyahp/prostate-cancer-dataset

GitHub. (2026). *GitHub Copilot* [AI coding assistant]. Visual Studio Code. https://github.com/features/copilot

Martin, G. (2025, May 7). *RStudio for beginners* [Video]. YouTube. https://www.youtube.com/watch?v=Kgwfuycn9_I

Posit Software, PBC. (2026). *RStudio User Guide: Get Started*. https://docs.posit.co/ide/user/ide/get-started/

R Graph Gallery. (n.d.). *Grouped, stacked and percent stacked barplot in ggplot2*. https://r-graph-gallery.com/48-grouped-barplot-with-ggplot2.html

Wickham, H., Çetinkaya-Rundel, M., & Grolemund, G. (2023). *R for Data Science* (2nd ed.). O'Reilly Media. https://r4ds.hadley.nz/

Wickham, H., Navarro, D., & Pedersen, T. L. (n.d.). *ggplot2 documentation*. https://ggplot2.tidyverse.org/

Posit Software, PBC. (n.d.). *Cheatsheets*. https://posit.co/resources/cheatsheets/

swirl. (n.d.). *Learn R, in R*. https://swirlstats.com/
