---
title: "DATA 607 Project 1: Code Base"
author: "Andre Thomson"
format:
  html:
    toc: true
    embed-resources: true
    code-fold: true
    code-summary: "Show code"
    code-tools: true
editor: visual
execute:
  warning: false
  message: false
---

## Approach

For this project, I will use the chess tournament text file from the course. It is not arranged like a normal CSV because each player uses two lines. The first line has the player's number, name, points, and round results. The second line has the state and pre-rating.

I will read the file into R and find the lines that start with a player number. I will split those lines at the vertical bars and collect the name, points, and opponent numbers. I will use the next line for the state and pre-rating.

Next, I will match each opponent number with that player's pre-rating and calculate the average. I will leave out byes and unplayed rounds because they do not have an opponent.

The final dataframe will contain the player's name, state, total points, pre-rating, and average opponent pre-rating. I will check that there are 64 players, look for missing or duplicate values, compare Gary Hua's result with the example, and export the table to CSV.

The code is folded in the HTML viewer. Click **Show code** when you want to see it.

## Setup

```{r}
# tidyverse: data wrangling, parsing, and charts
library(tidyverse)
# stringr: text matching and cleanup
library(stringr)
# knitr: clean HTML table output
library(knitr)
```

## Load the Tournament File

```{r}
# Look for the tournament file in common project locations first.
# This avoids using a path that only works on one computer.
local_files <- c(
  "tournamentinfo.txt",
  file.path("data", "tournamentinfo.txt"),
  file.path("Project 1", "tournamentinfo.txt")
)

input_file <- local_files[file.exists(local_files)][1]

# Online fallback: public copy of the same DATA 607 tournament dataset.
data_url <- paste0(
  "https://raw.githubusercontent.com/longflin/DATA-607/",
  "refs/heads/main/Project%201/tournamentinfo.txt"
)

if (!is.na(input_file)) {
  raw_lines <- readLines(input_file, warn = FALSE)
  data_source <- input_file
} else {
  raw_lines <- tryCatch(
    readLines(data_url, warn = FALSE),
    error = function(e) {
      stop(
        "Tournament data could not be loaded. ",
        "Put tournamentinfo.txt in the same folder as this QMD ",
        "or in a data/ folder, then render again. ",
        "The online fallback also could not be reached."
      )
    }
  )
  data_source <- data_url
}

cat("Data source:", data_source)
```

The code checks for a local copy first. If the file is not there, it tries an online copy of the same tournament dataset. This fixes the original working-directory problem while still allowing the course file to be used when it is available.

## Find the Player Records

```{r}
# Find lines that begin with a player number followed by a vertical bar
player_line_index <- which(
  str_detect(raw_lines, "^\\s*\\d+\\s*\\|")
)

length(player_line_index)
```

The first line for each player starts with a player number. I use that pattern to find the player records instead of depending on fixed row numbers.

## Parse the Player Information

```{r}
# Remove extra spaces around parsed text fields
clean_field <- function(x) {
  str_squish(x)
}

# Pull a 3- or 4-digit chess rating from the rating field
parse_rating <- function(x) {
  rating_match <- str_match(x, "(\\d{3,4})")[, 2]
  as.numeric(rating_match)
}

# Parse every player record and combine the results into one dataframe
players <- map_dfr(player_line_index, function(i) {

  line1 <- str_split(raw_lines[i], "\\|", simplify = TRUE)
  line2 <- str_split(raw_lines[i + 1], "\\|", simplify = TRUE)

  player_no <- as.integer(clean_field(line1[1]))
  player_name <- clean_field(line1[2])
  total_points <- parse_number(clean_field(line1[3]))

  state <- clean_field(line2[1])
  pre_rating <- parse_rating(clean_field(line2[2]))

  round_fields <- if (ncol(line1) >= 4) line1[1, 4:ncol(line1)] else character(0)

  opponent_numbers <- str_match(round_fields, "(\\d+)")[, 2] |>
    as.integer()

  tibble(
    player_no = player_no,
    name = player_name,
    state = state,
    total_points = total_points,
    pre_rating = pre_rating,
    opponent_numbers = list(opponent_numbers)
  )
})
```

This separates the fields from the two-line record and stores each player's opponent numbers for the next step.

## Check the Parsed Data

```{r}
dim(players)
head(players)
sum(is.na(players$name))
sum(is.na(players$pre_rating))
sum(duplicated(players$player_no))
```

The project description I am using expects 64 players. I also check for missing names, missing ratings, and duplicate player numbers before calculating opponent averages.

```{r}
stopifnot(nrow(players) == 64)
stopifnot(!anyDuplicated(players$player_no))
```

If one of these checks fails, the code stops instead of silently producing a bad result.

## Calculate Average Opponent Pre-Rating

```{r}
# Create a lookup table: player number -> pre-rating
rating_lookup <- players |>
  select(player_no, opponent_pre_rating = pre_rating)

# Expand opponent numbers, join opponent ratings, then average by player
final_data <- players |>
  select(
    player_no,
    name,
    state,
    total_points,
    pre_rating,
    opponent_numbers
  ) |>
  unnest_longer(
    opponent_numbers,
    values_to = "opponent_no",
    keep_empty = TRUE
  ) |>
  filter(!is.na(opponent_no)) |>
  left_join(
    rating_lookup,
    by = c("opponent_no" = "player_no")
  ) |>
  group_by(
    player_no,
    name,
    state,
    total_points,
    pre_rating
  ) |>
  summarise(
    average_opponent_pre_rating =
      round(mean(opponent_pre_rating, na.rm = TRUE)),
    .groups = "drop"
  ) |>
  arrange(player_no)
```

Byes and unplayed rounds do not produce an opponent number, so they are left out of the average.

## Final Table

```{r}
# Keep and rename only the five fields used in the final result
project1_table <- final_data |>
  transmute(
    Name = name,
    State = state,
    `Total Number of Points` = total_points,
    `Pre-Rating` = pre_rating,
    `Average Pre Tournament Chess Rating of Opponents` =
      average_opponent_pre_rating
  )

kable(
  project1_table,
  caption = "Project 1 Chess Tournament Results"
)
```

This is the table I will export for the project.

## Quality Checks

```{r}
nrow(project1_table)
colSums(is.na(project1_table))
sum(duplicated(project1_table$Name))
summary(project1_table)
```

I use these checks to make sure the final table has the expected number of rows and to catch missing or duplicate values.

### Gary Hua Check

```{r}
gary_hua_check <- project1_table |>
  filter(str_to_lower(Name) == "gary hua")

gary_hua_check
```

I included this check because the approach says I will compare Gary Hua's result with the course example. I will make that comparison from the rendered output rather than typing a result into the document before the code runs.

## Exploratory Charts

These charts are secondary to the required table. I kept them simple so they help me inspect the cleaned data without taking over the project.

### Pre-Rating Distribution

```{r}
ggplot(project1_table, aes(x = `Pre-Rating`)) +
  geom_histogram(
    bins = 12,
    fill = "steelblue",
    color = "white"
  ) +
  labs(
    title = "Distribution of Player Pre-Ratings",
    x = "Pre-Rating",
    y = "Number of Players"
  ) +
  theme_minimal()
```

The histogram gives a quick view of how the player ratings are distributed.

### Points by Player

```{r}
#| fig-height: 16
#| fig-width: 10

# Order players by points and use pre-rating to break ties.
# The taller figure gives all player names enough room to be readable.
points_plot <- project1_table |>
  arrange(`Total Number of Points`, `Pre-Rating`) |>
  mutate(Name = factor(Name, levels = Name))

ggplot(
  points_plot,
  aes(x = `Total Number of Points`, y = Name)
) +
  geom_segment(
    aes(
      x = 0,
      xend = `Total Number of Points`,
      y = Name,
      yend = Name
    ),
    color = "grey75",
    linewidth = 0.6
  ) +
  geom_point(
    size = 3,
    color = "steelblue"
  ) +
  labs(
    title = "Tournament Points by Player",
    subtitle = "Players are ordered by total tournament points",
    x = "Total Points",
    y = NULL
  ) +
  scale_x_continuous(
    breaks = seq(
      0,
      max(points_plot$`Total Number of Points`, na.rm = TRUE),
      by = 1
    )
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.y = element_text(size = 8),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )
```

The taller lollipop chart makes the player names readable. Players are ordered by total points, with pre-rating used to break ties. This chart is supporting analysis; the final project table remains the main output. \### Player Rating and Opponent Rating

```{r}
ggplot(
  project1_table,
  aes(
    x = `Pre-Rating`,
    y = `Average Pre Tournament Chess Rating of Opponents`
  )
) +
  geom_point(
    alpha = 0.75,
    color = "steelblue"
  ) +
  labs(
    title = "Player Rating vs. Average Opponent Rating",
    x = "Player Pre-Rating",
    y = "Average Opponent Pre-Rating"
  ) +
  theme_minimal()
```

This scatterplot compares two numeric variables from the final dataset.

## Export the CSV

```{r}
# Save the final table as a CSV file
write_csv(
  project1_table,
  "project1_chess_results.csv"
)
```

The exported CSV contains only the five requested result columns.

## AI Use

I used GitHub Copilot in Visual Studio Code to help review parts of the R code, troubleshoot errors, and improve code comments. I reviewed the suggestions, kept the code I understood, and checked the results when running the project. The analysis and final results come from the R code and tournament data used in this project.

## References

DATA 607 chess tournament dataset (public copy used only as a fallback):\
https://raw.githubusercontent.com/longflin/DATA-607/refs/heads/main/Project%201/tournamentinfo.txt

Wickham, H., Cetinkaya-Rundel, M., & Grolemund, G. *R for Data Science (2e).*

R Graph Gallery. Lollipop plot examples.

R Graph Gallery. Histogram examples.
